Estimating process-based model parameters from species distribution data using the evolutionary algorithm CMA-ES

### 2023-03-28 manuscript/paper.Rmd
small changes before resubmission (minor revision)

---
output: 
  stevetemplates::article:
    keep_tex: true
    fig_caption: true
    number_sections: yes
bibliography: bib/paper.bib
csl: bib/mee.csl
title: "A novel approach to inverse calibration of process-based models using evolutionary algorithm and species distribution data"
author:
- name: Victor Van der Meersch \footnote{Corresponding author - victor.vandermeersch@cefe.cnrs.fr}, Isabelle Chuine
  affiliation: CEFE, Université de Montpellier, CNRS, EPHE, IRD, Montpellier, France
abstract: "1. Two main types of species distribution models are used to project species range shifts in future climatic conditions: correlative and process-based models. Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and their calibration methods (dependent *vs* generally independent of the species observed distributions). \n\n 2. One of the limitations to the use of process-based models is the difficulty to parameterize them for a large number of species, while one of the advantages of correlative SDMs is their easier calibration using solely species occurrences}. We investigated the feasibility of using an evolutionary algorithm (called covariance matrix adaptation evolution strategy, CMA-ES) to calibrate process-based models using species distribution data. This method is well established in some fields (robotics, aerospace research, ...), but has never been used, to our knowledge, in ecology, despite its ability to deal with very large space dimensions. Using tree species occurrence data across Europe, we adapted the CMA-ES algorithm to find appropriate values of model parameters. We estimated simultaneously 27 to 77 parameters of two process-based models simulating forest tree's ecophysiology for three species with varying range sizes and geographical distributions. \n\n 3. CMA-ES provided parameter estimates leading to better prediction of species distribution than parameter estimates based on experts knowledge. Our results also revealed that some model parameters and processes were strongly dependent, and different parameter combinations could therefore lead to high model accuracy. \n\n 4. \\textcolor{darkgreen}{We conclude that} CMA-ES is an efficient state-of-the-art method to calibrate process-based models with a large number of parameters using species occurrence data. Inverse modelling using CMA-ES is a powerful method to calibrate process-based parameters which can hardly be measured. However, the method does not warranty that parameter estimates are correct because of several sources of bias, similarly to correlative models, and expert knowledge is required to validate results. \n\n"
keywords: "calibration, evolutionary algorithm, cma-es, species distribution model, process-based model, trees"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
spacing: double
endnote: no
biblatex: true
link-citations: true

@@

Calibration (i.e. parameter setting and estimation) is a fundamental step in the modelling process. It allows the model to reproduce the reality with more or less success. The result of the calibration provides insights on the ability of the model to reproduce and explain reality (model predictive power). Calibration of complex models such as process-based SDMs is time-consuming, and modelers are often challenged by the dimension of the parameter space, the complexity of the possible correlations among parameters, and the scarcity of observed experimental data to calibrate them. Therefore, although the philosophy of such models is to measure parameters, statistical inference might be useful when data are not yet available to infer parameter values. Parameter inference can be achieved through several methods which have been developed in the last decades. \textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration. On the one hand, statistical calibration assumes a data-generating model, and the likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically. This likelihood is the foundation of maximum likelihood estimation and Bayesian inference, two commonly used methods for parameter estimation. In practice, deriving the likelihood function can be quite challenging for complex models with many parameters. One approach is to use numerical methods such as simulation-based inference or Approximate Bayesian Computation} [e.g. @Hartig2014]\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, using an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative, even though it may not have the same statistical rigor. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data. This optimization step can be achieved with many methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}     
Recently, an approach belonging to the evolutionary algorithm family, called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has been proposed [@Hansen2001]. One of the advantages of CMA-ES is its ability to cumulate information over iterations in order to adapt its own parameters (in particular the covariance matrix), which makes it more robust to noise. CMA-ES is especially performant for non-separable problems (i.e. when the model parameters are dependent) and large search space. This method has been successfully applied in various fields such as aerospace [e.g. @Collange2010], optics [e.g. @Gagne2008], and robotics [e.g. @Hill2020]. CMA-ES is acknowledged to be one of the most efficient approaches in continuous black-box optimization [@Hansen2010] but to our knowledge has never been used in ecology.

@@

\textcolor{darkgreen}{Our results showed that CMA-ES is a feasible and efficient optimizer for inverse calibration of complex ecological models. The algorithm was able to find parameter sets that significantly improved the predictions of the calibrated models compared to the expert parametrization.  However, our study also highlighted the issue of non-identifiability of parameter values due to the data limitation and the dependence between process-based model parameters, which may result in diverging parameter values even though the calibrated models describe the observed species distribution well. Nonetheless, our study provides a promising methodological framework for calibrating process-based models using CMA-ES, and for spreading the use of such models in predictive ecology.}

@@

CMA-ES is a "generic" optimizer which can be applied to various problems. It is easy to use as it does not require an extensive tuning to  efficiently explore the parameter space. We only had to choose the number of candidate solutions $\lambda$, and the initial search region (initial starting point and step size $\sigma$). As well as being quasi parameter-free, CMA-ES has several structural advantages particularly useful for complex optimization problems. \textcolor{darkgreen}{Firstly, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters. Secondly, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }[@Hansen2011]\textcolor{darkgreen}{. Lastly, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence }[@Hansen2001]\textcolor{darkgreen}{, allowing the algorithm to explore more of the search space.}
CMA-ES has been shown to outperform several other optimization algorithms [@Hansen2010], and is usually the most efficient method when the target cost (i.e. the number of objective function evaluations) is about $100*N$ [$N$ being the dimension of the parameter search space, @Baeck2013].

@@

Our goal here was to investigate the performance of CMA-ES to calibrate quite complex process-based species distribution models using species occurrence data. We did not attempt to validate our parametrizations using temporally or spatially independent data, and AUC was only used to determine if model outputs were consistent with species distributions. However, AUC is scale-invariant (it measures how well predictions are ranked rather than their absolute values), and prevents us from obtaining calibrated outputs with more ecological meaning. For example, in PHENOFIT, one species with a fitness of $0.8$ could be considered as absent whereas another one as present. Further work could thus be conducted to examine the effects of choosing a different objective function.

@@

ERA5-Land dataset is available on the [Copernicus Climate Change Service website](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-land?tab=overview). EU-SoilHydroGrids is available on the 
[European Soil Data Centre website](https://esdac.jrc.ec.europa.eu/content/3d-soil-hydraulic-database-europe-1-km-and-250-m-resolution). SoilGrids250m is available on the [International Soil Reference and Information Centre website](https://www.isric.org/explore/soilgrids). EU-Forest database is available on [FigShare](https://figshare.com/collections/A_high-resolution_pan-European_tree_occurrence_dataset/3288407). The R code associated with this work will be made available on GitLab upon acceptance of the manuscript.

>>>>

---
output: 
  stevetemplates::article:
    keep_tex: true
    fig_caption: true
    number_sections: yes
bibliography: bib/paper.bib
csl: bib/mee.csl
title: "Estimating process-based model parameters from species distribution data using the evolutionary algorithm CMA-ES"
author:
- name: Victor Van der Meersch \footnote{Corresponding author - victor.vandermeersch@cefe.cnrs.fr}, Isabelle Chuine
  affiliation: CEFE, Université de Montpellier, CNRS, EPHE, IRD, Montpellier, France
abstract: "1. Two main types of species distribution models are used to project species range shifts in future climatic conditions: correlative and process-based models. Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and and their calibration methods \\textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}. \n\n 2. One of the limitations to the use of process-based models is the difficulty to parameterize them for a large number of species \\textcolor{darkgreen}{compared to correlative SDMs}. We investigated the feasibility of using an evolutionary algorithm (called covariance matrix adaptation evolution strategy, CMA-ES) to calibrate process-based models using species distribution data. This method is well established in some fields (robotics, aerospace research, ...), but has never been used, to our knowledge, in ecology, despite its ability to deal with very large space dimensions. Using tree species occurrence data across Europe, we adapted the CMA-ES algorithm to find appropriate values of model parameters. We estimated simultaneously 27 to 77 parameters of two process-based models simulating forest tree's ecophysiology for three species with varying range sizes and geographical distributions. \n\n 3. CMA-ES provided parameter estimates leading to better prediction of species distribution than parameter estimates based on experts knowledge. Our results also revealed that some model parameters and processes were strongly dependent, and different parameter combinations could therefore lead to high model accuracy. \n\n 4. \\textcolor{darkgreen}{We conclude that} CMA-ES is an efficient state-of-the-art method to calibrate process-based models with a large number of parameters using species occurrence data. Inverse modelling using CMA-ES is a powerful method to calibrate process-based parameters which can hardly be measured. However, the method does not warranty that parameter estimates are correct because of several sources of bias, similarly to correlative models, and expert knowledge is required to validate results. \n\n"
keywords: "calibration, evolutionary algorithm, cma-es, species distribution model, process-based model, trees"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
spacing: double
endnote: no
biblatex: true
link-citations: true

@@

Calibration (i.e. parameter setting and estimation) is a fundamental step in the modelling process. It allows the model to reproduce the reality with more or less success. The result of the calibration provides insights on the ability of the model to reproduce and explain reality (model predictive power). Calibration of complex models such as process-based SDMs is time-consuming, and modelers are often challenged by the dimension of the parameter space, the complexity of the possible correlations among parameters, and the scarcity of observed experimental data to calibrate them. Therefore, although the philosophy of such models is to measure parameters, statistical inference might be useful when data are not yet available to infer parameter values. Parameter inference can be achieved through several methods which have been developed in the last decades. \textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration. On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically. This likelihood is the foundation of *maximum likelihood estimation* and *Bayesian inference*, two commonly used methods for parameter estimation. In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding. One approach is to use numerical methods such as *simulation-based inference* or *approximate Bayesian computation*} [e.g. @Hartig2014]\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration. This optimization step can be achieved with several methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}     
Recently, an approach belonging to the evolutionary algorithm family, called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has been proposed [@Hansen2001]. One of the advantages of CMA-ES is its ability to cumulate information over iterations in order to adapt its own parameters (in particular the covariance matrix), which makes it more robust to noise. CMA-ES is especially performant for non-separable problems (i.e. when the model parameters are dependent) and large search space. This method has been successfully applied in various fields such as aerospace [e.g. @Collange2010], optics [e.g. @Gagne2008], and robotics [e.g. @Hill2020]. CMA-ES is acknowledged to be one of the most efficient approaches in continuous black-box optimization [@Hansen2010] but to our knowledge has never been used in ecology.

@@

\textcolor{darkgreen}{Our results showed that CMA-ES is an efficient optimizer for inverse calibration of complex ecological models. The algorithm was able to find parameter sets that significantly improved the predictions of the calibrated models compared to the expert parametrization.  However, our study also highlighted the issue of non-identifiability of parameter values due to the data limitation and the dependence between process-based model parameters, which may result in diverging parameter values even though the calibrated models describe the observed species distribution well.}

@@

CMA-ES is a "generic" optimizer which can be applied to various problems. It is easy to use as it does not require an extensive tuning to  efficiently explore the parameter space. We only had to choose the number of candidate solutions $\lambda$, and the initial search region (initial starting point and step size $\sigma$). As well as being quasi parameter-free, CMA-ES has several structural advantages particularly useful for complex optimization problems. \textcolor{darkgreen}{First, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters. Second, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }[@Hansen2011]\textcolor{darkgreen}{. Last, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence }[@Hansen2001]\textcolor{darkgreen}{, allowing the algorithm to explore more of the search space.}
CMA-ES has been shown to outperform several other optimization algorithms [@Hansen2010], and is usually the most efficient method when the target cost (i.e. the number of objective function evaluations) is about $100*N$ [$N$ being the dimension of the parameter search space, @Baeck2013].

@@

Our goal here was to investigate the performance of CMA-ES to calibrate quite complex process-based species distribution models using species occurrence data. We did not attempt to validate our parametrizations using temporally or spatially independent data, and AUC was only used to determine if model outputs were consistent with species distributions. \textcolor{darkgreen}{However, AUC is scale-invariant: it measures how well predictions are ranked rather than their absolute values. For example, with PHENOFIT, a species with a fitness of $0.8$ could be considered as absent while another one with the same fitness could be considered as present. Therefore, when it is used as an objective function for model calibration, we probably lack some precision and consistency among species’ parameters estimation.} Further work could thus be conducted to examine the effects of choosing a different objective function.

@@

ERA5-Land dataset is available on the [Copernicus Climate Change Service website](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-land?tab=overview). EU-SoilHydroGrids is available on the 
[European Soil Data Centre website](https://esdac.jrc.ec.europa.eu/content/3d-soil-hydraulic-database-europe-1-km-and-250-m-resolution). SoilGrids250m is available on the [International Soil Reference and Information Centre website](https://www.isric.org/explore/soilgrids). EU-Forest database is available on [FigShare](https://figshare.com/collections/A_high-resolution_pan-European_tree_occurrence_dataset/3288407). The R code associated with this work is available on [this GitHub repository](https://github.com/vvandermeersch/inverse_calibration), as well as on Zenodo ([10.5281/zenodo.7774981](https://doi.org/10.5281/zenodo.7774981)).

### 2023-03-28 manuscript/paper.tex
small changes before resubmission (minor revision)

\title{A novel approach to inverse calibration of process-based models
using evolutionary algorithm and species distribution data  }

@@

\noindent \begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Two main types of species distribution models are used to project
  species range shifts in future climatic conditions: correlative and
  process-based models. Although there is some continuity between these
  two types of models, they are fundamentally different in their
  hypotheses (statistical relationships \emph{vs} mechanistic
  relationships) and their calibration methods (dependent \emph{vs}
  generally independent of the species observed distributions).
\item
  One of the limitations to the use of process-based models is the
  difficulty to parameterize them for a large number of species, while
  one of the advantages of correlative SDMs is their easier calibration
  using solely species occurrences\}. We investigated the feasibility of
  using an evolutionary algorithm (called covariance matrix adaptation
  evolution strategy, CMA-ES) to calibrate process-based models using
  species distribution data. This method is well established in some
  fields (robotics, aerospace research, \ldots), but has never been
  used, to our knowledge, in ecology, despite its ability to deal with
  very large space dimensions. Using tree species occurrence data across
  Europe, we adapted the CMA-ES algorithm to find appropriate values of
  model parameters. We estimated simultaneously 27 to 77 parameters of
  two process-based models simulating forest tree's ecophysiology for
  three species with varying range sizes and geographical distributions.
\item
  CMA-ES provided parameter estimates leading to better prediction of
  species distribution than parameter estimates based on experts
  knowledge. Our results also revealed that some model parameters and
  processes were strongly dependent, and different parameter
  combinations could therefore lead to high model accuracy.
\item
  \textcolor{darkgreen}{We conclude that} CMA-ES is an efficient
  state-of-the-art method to calibrate process-based models with a large
  number of parameters using species occurrence data. Inverse modelling
  using CMA-ES is a powerful method to calibrate process-based
  parameters which can hardly be measured. However, the method does not
  warranty that parameter estimates are correct because of several
  sources of bias, similarly to correlative models, and expert knowledge
  is required to validate results.
\end{enumerate}

@@

Calibration (i.e.~parameter setting and estimation) is a fundamental
step in the modelling process. It allows the model to reproduce the
reality with more or less success. The result of the calibration
provides insights on the ability of the model to reproduce and explain
reality (model predictive power). Calibration of complex models such as
process-based SDMs is time-consuming, and modelers are often challenged
by the dimension of the parameter space, the complexity of the possible
correlations among parameters, and the scarcity of observed experimental
data to calibrate them. Therefore, although the philosophy of such
models is to measure parameters, statistical inference might be useful
when data are not yet available to infer parameter values. Parameter
inference can be achieved through several methods which have been
developed in the last decades.
\textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration. On the one hand, statistical calibration assumes a data-generating model, and the likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically. This likelihood is the foundation of maximum likelihood estimation and Bayesian inference, two commonly used methods for parameter estimation. In practice, deriving the likelihood function can be quite challenging for complex models with many parameters. One approach is to use numerical methods such as simulation-based inference or Approximate Bayesian Computation}
(e.g. \protect\hyperlink{ref-Hartig2014}{Hartig \emph{et al.}
2014})\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, using an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative, even though it may not have the same statistical rigor. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data. This optimization step can be achieved with many methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}\\
Recently, an approach belonging to the evolutionary algorithm family,
called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has
been proposed (\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier
2001}). One of the advantages of CMA-ES is its ability to cumulate
information over iterations in order to adapt its own parameters (in
particular the covariance matrix), which makes it more robust to noise.
CMA-ES is especially performant for non-separable problems (i.e.~when
the model parameters are dependent) and large search space. This method
has been successfully applied in various fields such as aerospace (e.g.
\protect\hyperlink{ref-Collange2010}{Collange \emph{et al.} 2010}),
optics (e.g. \protect\hyperlink{ref-Gagne2008}{Gagné \emph{et al.}
2008}), and robotics (e.g. \protect\hyperlink{ref-Hill2020}{Hill
\emph{et al.} 2020}). CMA-ES is acknowledged to be one of the most
efficient approaches in continuous black-box optimization
(\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.} 2010}) but to
our knowledge has never been used in ecology.

@@

CMA-ES is a ``generic'' optimizer which can be applied to various
problems. It is easy to use as it does not require an extensive tuning
to efficiently explore the parameter space. We only had to choose the
number of candidate solutions \(\lambda\), and the initial search region
(initial starting point and step size \(\sigma\)). As well as being
quasi parameter-free, CMA-ES has several structural advantages
particularly useful for complex optimization problems.
\textcolor{darkgreen}{Firstly, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters. Secondly, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }(\protect\hyperlink{ref-Hansen2011}{Hansen
\emph{et al.}
2011})\textcolor{darkgreen}{. Lastly, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence }(\protect\hyperlink{ref-Hansen2001}{Hansen
\& Ostermeier
2001})\textcolor{darkgreen}{, allowing the algorithm to explore more of the search space.}
CMA-ES has been shown to outperform several other optimization
algorithms (\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.}
2010}), and is usually the most efficient method when the target cost
(i.e.~the number of objective function evaluations) is about \(100*N\)
(\(N\) being the dimension of the parameter search space,
\protect\hyperlink{ref-Baeck2013}{Bäck \emph{et al.} 2013}).

@@

Our goal here was to investigate the performance of CMA-ES to calibrate
quite complex process-based species distribution models using species
occurrence data. We did not attempt to validate our parametrizations
using temporally or spatially independent data, and AUC was only used to
determine if model outputs were consistent with species distributions.
However, AUC is scale-invariant (it measures how well predictions are
ranked rather than their absolute values), and prevents us from
obtaining calibrated outputs with more ecological meaning. For example,
in PHENOFIT, one species with a fitness of \(0.8\) could be considered
as absent whereas another one as present. Further work could thus be
conducted to examine the effects of choosing a different objective
function.

@@

ERA5-Land dataset is available on the
\href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-land?tab=overview}{Copernicus
Climate Change Service website}. EU-SoilHydroGrids is available on the
\href{https://esdac.jrc.ec.europa.eu/content/3d-soil-hydraulic-database-europe-1-km-and-250-m-resolution}{European
Soil Data Centre website}. SoilGrids250m is available on the
\href{https://www.isric.org/explore/soilgrids}{International Soil
Reference and Information Centre website}. EU-Forest database is
available on
\href{https://figshare.com/collections/A_high-resolution_pan-European_tree_occurrence_dataset/3288407}{FigShare}.
The R code associated with this work will be made available on GitLab
upon acceptance of the manuscript.

>>>>

\title{Estimating process-based model parameters from species
distribution data using the evolutionary algorithm CMA-ES  }

@@

\noindent \begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Two main types of species distribution models are used to project
  species range shifts in future climatic conditions: correlative and
  process-based models. Although there is some continuity between these
  two types of models, they are fundamentally different in their
  hypotheses (statistical relationships \emph{vs} mechanistic
  relationships) and and their calibration methods
  \textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}.
\item
  One of the limitations to the use of process-based models is the
  difficulty to parameterize them for a large number of species
  \textcolor{darkgreen}{compared to correlative SDMs}. We investigated
  the feasibility of using an evolutionary algorithm (called covariance
  matrix adaptation evolution strategy, CMA-ES) to calibrate
  process-based models using species distribution data. This method is
  well established in some fields (robotics, aerospace research,
  \ldots), but has never been used, to our knowledge, in ecology,
  despite its ability to deal with very large space dimensions. Using
  tree species occurrence data across Europe, we adapted the CMA-ES
  algorithm to find appropriate values of model parameters. We estimated
  simultaneously 27 to 77 parameters of two process-based models
  simulating forest tree's ecophysiology for three species with varying
  range sizes and geographical distributions.
\item
  CMA-ES provided parameter estimates leading to better prediction of
  species distribution than parameter estimates based on experts
  knowledge. Our results also revealed that some model parameters and
  processes were strongly dependent, and different parameter
  combinations could therefore lead to high model accuracy.
\item
  \textcolor{darkgreen}{We conclude that} CMA-ES is an efficient
  state-of-the-art method to calibrate process-based models with a large
  number of parameters using species occurrence data. Inverse modelling
  using CMA-ES is a powerful method to calibrate process-based
  parameters which can hardly be measured. However, the method does not
  warranty that parameter estimates are correct because of several
  sources of bias, similarly to correlative models, and expert knowledge
  is required to validate results.
\end{enumerate}

@@

Calibration (i.e.~parameter setting and estimation) is a fundamental
step in the modelling process. It allows the model to reproduce the
reality with more or less success. The result of the calibration
provides insights on the ability of the model to reproduce and explain
reality (model predictive power). Calibration of complex models such as
process-based SDMs is time-consuming, and modelers are often challenged
by the dimension of the parameter space, the complexity of the possible
correlations among parameters, and the scarcity of observed experimental
data to calibrate them. Therefore, although the philosophy of such
models is to measure parameters, statistical inference might be useful
when data are not yet available to infer parameter values. Parameter
inference can be achieved through several methods which have been
developed in the last decades.
\textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration. On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically. This likelihood is the foundation of *maximum likelihood estimation* and *Bayesian inference*, two commonly used methods for parameter estimation. In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding. One approach is to use numerical methods such as *simulation-based inference* or *approximate Bayesian computation*}
(e.g. \protect\hyperlink{ref-Hartig2014}{Hartig \emph{et al.}
2014})\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration. This optimization step can be achieved with several methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}\\
Recently, an approach belonging to the evolutionary algorithm family,
called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has
been proposed (\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier
2001}). One of the advantages of CMA-ES is its ability to cumulate
information over iterations in order to adapt its own parameters (in
particular the covariance matrix), which makes it more robust to noise.
CMA-ES is especially performant for non-separable problems (i.e.~when
the model parameters are dependent) and large search space. This method
has been successfully applied in various fields such as aerospace (e.g.
\protect\hyperlink{ref-Collange2010}{Collange \emph{et al.} 2010}),
optics (e.g. \protect\hyperlink{ref-Gagne2008}{Gagné \emph{et al.}
2008}), and robotics (e.g. \protect\hyperlink{ref-Hill2020}{Hill
\emph{et al.} 2020}). CMA-ES is acknowledged to be one of the most
efficient approaches in continuous black-box optimization
(\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.} 2010}) but to
our knowledge has never been used in ecology.

@@

CMA-ES is a ``generic'' optimizer which can be applied to various
problems. It is easy to use as it does not require an extensive tuning
to efficiently explore the parameter space. We only had to choose the
number of candidate solutions \(\lambda\), and the initial search region
(initial starting point and step size \(\sigma\)). As well as being
quasi parameter-free, CMA-ES has several structural advantages
particularly useful for complex optimization problems.
\textcolor{darkgreen}{First, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters. Second, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }(\protect\hyperlink{ref-Hansen2011}{Hansen
\emph{et al.}
2011})\textcolor{darkgreen}{. Last, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence }(\protect\hyperlink{ref-Hansen2001}{Hansen
\& Ostermeier
2001})\textcolor{darkgreen}{, allowing the algorithm to explore more of the search space.}
CMA-ES has been shown to outperform several other optimization
algorithms (\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.}
2010}), and is usually the most efficient method when the target cost
(i.e.~the number of objective function evaluations) is about \(100*N\)
(\(N\) being the dimension of the parameter search space,
\protect\hyperlink{ref-Baeck2013}{Bäck \emph{et al.} 2013}).

@@

Our goal here was to investigate the performance of CMA-ES to calibrate
quite complex process-based species distribution models using species
occurrence data. We did not attempt to validate our parametrizations
using temporally or spatially independent data, and AUC was only used to
determine if model outputs were consistent with species distributions.
\textcolor{darkgreen}{However, AUC is scale-invariant: it measures how well predictions are ranked rather than their absolute values. For example, with PHENOFIT, a species with a fitness of $0.8$ could be considered as absent while another one with the same fitness could be considered as present. Therefore, when it is used as an objective function for model calibration, we probably lack some precision and consistency among species’ parameters estimation.}
Further work could thus be conducted to examine the effects of choosing
a different objective function.

@@

ERA5-Land dataset is available on the
\href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-land?tab=overview}{Copernicus
Climate Change Service website}. EU-SoilHydroGrids is available on the
\href{https://esdac.jrc.ec.europa.eu/content/3d-soil-hydraulic-database-europe-1-km-and-250-m-resolution}{European
Soil Data Centre website}. SoilGrids250m is available on the
\href{https://www.isric.org/explore/soilgrids}{International Soil
Reference and Information Centre website}. EU-Forest database is
available on
\href{https://figshare.com/collections/A_high-resolution_pan-European_tree_occurrence_dataset/3288407}{FigShare}.
The R code associated with this work is available on
\href{https://github.com/vvandermeersch/inverse_calibration}{this GitHub
repository}, as well as on Zenodo
(\href{https://doi.org/10.5281/zenodo.7774981}{10.5281/zenodo.7774981}).
